Locality and Computational Reliability of Linear Response Calculations for Molecular Systems

### 2018-07-11 AnalyticStructure_main.tex
Update on Overleaf.

%\title{On the analytic structure of the linear response susceptibility in open systems}
\title{Locality and Computational Reliability of Linear Response Calculations for Molecular Systems}% Localization properties, observable character and computational reliability of the linear response in open systems}
\author{Marco D'Alessandro}%
\affiliation{Istituto di Struttura della Materia-CNR (ISM-CNR), Via del Fosso del Cavaliere 100, 00133 Roma, Italia}
\author{Luigi Genovese}
\affiliation{Laboratoire de Simulation Atomistique (LSim), SP2M, INAC, CEA-UJF, 17 Av. des Martyrs,
38054 Grenoble, France}
\date{\today}

@@

\section{Introduction}
Since the days of their foundations, disciplines like Computational Physics or Quantum Chemistry always had to deal with the problem of the 
computational reliability of the results. More specifically, a computational result may be considered as reliable for various reasons.
It might be, for example, in agreement with some previous observation or measurement that the employed computational model is susceptible to reproduce.
The model itself may in this way be classified as ``accurate''; it can then be employed to extract, \textit{in silico},
other quantities that might be then used to \emph{predict} or \emph{interpret} experiments.
Another concept of reliability is the concept of  ``precision'', which is more focused to the actual results
that a given computational model provides for a specific set of quantities, regardless of their potential agreement with 
experimental data. The aim of precise computational approach is to reduce the \emph{computational} uncertainties
of quantities extracted out of a well-defined model, and provide reference results, to which other computer codes
employing the same model can be compared.

@@

The problem is even more stringent for calculations that refer to quantities \emph{beyond} ground state, like
for instance the time-dependent density functional theory (TDDFT) \cite{casida1995,runge1984,onida2002},
from which quantities like optical excitations or polarizability tensors
of a given system might be studied and put in relation with experimental data.
Here also, a plethora of computational approaches exist, that employ different numerical formalism and
basis sets, and it will be of great importance to identify quantities that can be ``measured'' in a given model, to assess
and quantify their computational uncertainty.

In this paper, we focus our attention to Linear-Response treatments of \emph{molecular} systems, 
in particular referring to LR-TDDFT calculation. We will investigate the impact that 
\emph{open} (i.e. Isolated) Boundary Conditions (BC) that have to be imposed to the system will have on the computational evaluation of LR quantities.
In order to do this, we perform analytic considerations on the equations beyond LR treatment.
Such considerations will enable us to distinguish quantities that \textit{a priori} can be compared among different treatments - and therefore tested for reproducibility - 
from others that would \emph{explicitly} depend of the the employed computational treatment. 
In particular, we will discuss the interplay between reproducible LR quantities and the \emph{locality} of the related objects.
% In order to do this, we perform analytic considerations on the equations beyond LR treatment,
% which will help us identifying how to distinguish -- \textit{a priori} -- quantities that are susceptible to be 
% compared among different treatments, as they are \emph{intrinsically} associated to localized objects,
% from quantities that \emph{explicitly} depend of the the employed computational treatment.

The paper is organized as follows.
Initially we will inspect how the question of reproducibility may be expressed 
for ground state calculations of molecules. 
Following these guidelines, we will then move to LR equations,
by first considering the equations of motion of the response density operator, given
the application of a perturbing field.
Then we will consider the behaviour of the ``free oscillations'', i.e. the excited states
of the molecule.
Our considerations will also reveal useful for the understanding of the \emph{analytic structure} of LR quantities like
the linear susceptibility of the system (i.e. the reducible polarizability).
We support our considerations with numerical results on simple LR-TDDFT quantities.
For readability purposes, the details of each calculation are explained in the appendix.

Although we refer, as anticipated, to TDDFT models, such considerations might also be transposed to Many-Body Perturbation Theory
approaches, in the context Quasi-Particle equations.

@@

On the other hand, positive energy solutions behave differently. 
Even though the potential is localized, their positive energy makes them behave 
as spherical waves and its convolution with the source term  gives raise to a delocalized wavefunction $\psi(\r)$ 
in the whole accessible space.
As they cannot be normalized they belong to the essential spectrum of the Hamiltonian operator and form a \emph{continuum} of states.
Thanks to the Lippmann-Schwinger relation of Eq.~\eqref{SEHelmholtz1} they can be put in bijection with the eigenstates of the 
free-particle Hamiltonian $\op H_s$. We may say that their quantum numbers are determined by the Laplacian operator, 
regardless of the particular features of the potential.

It is easy to see from these concepts that a computational treatment that is tailored for the discretization of bound states
will not be effective for continuum states, due to the different long-range asymptotic behavior - which in turn is a clear consequence of the 
BC of the problem. In a numerical treatment, discretizing the problem with a \emph{finite}
number of degrees of freedom, we will only have access to a \emph{pseudo}-continuum of states,
whose energy values and density of states will depend on the representation of the kinetic operator
in the basis set. As a consequence, numerical eigenstates with $\epsilon > 0$ will have a value of their energy that will \emph{always}
depend on the computational setup.

A direct numerical evidence of this problem has been discussed in \cite{boffi2016}. The authors of this paper analyze the effects of a finite basis set on the 
occupied and virtual orbitals of a molecular system and establish a clear correspondence between the orbital localization character, 
and the independence of its energy with respect to the basis set. 
% It is shown that (localized) orbitals with a bound-state behavior in the long range constitute a genuine discrete set and a finite basis is 
% able to provide a reliable description of these states if the sampled domain contains their natural sizes. On the contrary, unbound (delocalized) orbitals form a continuum of states in the 
% true system with open boundaries. For these class of objects the presence of a finite computational domain has a deeper impact since introduces a fictitious discretization of the energy levels. 
On the contrary, it is clearly shown that the energies of the unbound orbitals are not stable with respect to a change of the size of the computational domain and approach to zero as long as it is increased.

\section{Fluctuation states. A localization argument for the response density}
It is evident that the arguments presented in the above section have a limited interest for the study
of Ground State quantities of molecules. By definition, all these quantities can be expressed as a 
functional which \emph{entirely} depends on bound states. As an illustration it is sufficient to recall
the Hohenberg-Kohn theorem that states that the Ground-State (GS) energy is a functional of the GS
charge density $\dmnot$. For molecular systems that are electronically stable (i.e. no anionic or metastable configurations),
such quantity is localized, in the sense of the above section. Stated otherwise, 
\emph{there is no need to express efficiently any portion of the continuum spectrum} to have a reliable GS treatment of molecules.

@@

The response density satisfies an equation of motion written in the form of a quantum Liouville operator (for example see \cite{baroni2008})
\be\lb{LiouvillianRhopomegaDef1}
\left(\omega - \Liouv\right) \dm'_\Phi(\omega) =  \commutator{\op\Phi(\omega)}{\dmnot} \;,
\ee
where $\op\Phi(\omega)$ represents the perturbing field. 
The Liouvillian superoperator $\Liouv$ for self-consistent systems is expressed as the sum of the unperturbed part plus a coupling term. The action of these 
terms reads, respectively
\be\lb{LiouZeroDef1}
\Liouv \op O \equiv \left(\Liouvnot + \coupl \right) \op O = \commutator{\hnot}{\op O} +\commutator{\op V'[\op O]}{\dmnot}  \;,
\ee
where $\op V'[\op O] \equiv \int \dd \r \dd \r'\frac{\delta \op V[\dmnot]}{\delta \rho(\r,\r')} O(\r,\r')$.

The inspection of the linear order time evolution of the eigenstates of $\hnot$ evidence that the response density acts as a transition 
operator, linking the occupied and empty subspaces of $\hnot$, and satisfies the \emph{transverse} condition  
\be\lb{RhopTransverseDef1}
\dm'_\Phi = \dm_{\perp} \doteq \dmnot\dm'_\Phi\op Q_0 + \op Q_0\dm'_\Phi\dmnot \;,
\ee
where $\op Q_0=\identity-\dmnot$ is the projector in the subspace orthogonal to the occupied states of $\hnot$. 
Due to this fact we can introduce an explicit representation of the 
response density parametrized as
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'_\Phi(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p^\Phi(-\omega)} + \ketbra{f_p^\Phi(\omega)}{\psi_p}\right) \;.
\ee
Here we have introduced the set of $\omega$-dependent \emph{fluctuation states} 
$\ket{f_p^\Phi(\omega)} = \dm'_\Phi(\omega) \ket{\psi_p}$, that have values in the unoccupied subspace of $\hnot$. 
An analysis of the Liouville equation \eqref{LiouvillianRhopomegaDef1} for the response density written in this fashion evidences that the equations of motion of fluctuation states are 
written as a modified Sternheimer equation~\cite{mahan1980,giustino2012,giustino2014}
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p^\Phi(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'[\dm'_\Phi](\omega))\ket{\psi_p} \;,
\ee
and we observe that the first-order Hamiltonian contains, 
apart from the perturbing field, a further term $\op V'[\dm'_\Phi]$ due to the density-dependence of $\hnot$.
% The interesting aspect of these equations in the perspective of the present analysis is that they allow us to present a localization argument for the fluctuation states using a procedure  
% analogous to the one depicted in section \ref{SEopenSystem}. So we split the unperturbed Hamiltonian as the sum of the kinetic $\op T$ and potential $\op V$ term and express the solution of 
% \eqref{fluctuationStateEqMotion1} in the form of a self-consistent inhomogeneous Helmholtz equation
It is easy to see that the solution of equation \eqref{fluctuationStateEqMotion1} can be casted in the form of
a set of scattering problems:
%a self-consistent inhomogeneous Helmholtz equation
\be\lb{rhoPrimeFluctuationStateDef2}
\ket{f_p^\Phi(\omega)} = \op G_H(\omega+\epsilon_p)\left(\op V\ket{f_p^\Phi(\omega)} + \ket{s_p^\Phi(\omega)} \right)\;,
\ee
with%:here in this case the source term contains further addend apart from the unperturbed potential and reads:
$%\be
\ket{s_p^\Phi(\omega)} = \op Q_0(\op V'[\dm'_\Phi](\omega) + \op \Phi(\omega))\ket{\psi_p}
$. %\ee

It can be seen that for each choice of $p$, the value $\omega = |\eps_p|$ represents the threshold level 
that govern the localization properties of the corresponding fluctuation state.
As a \emph{stringent} hypothesis, let us now assume that the ket $\ket{s_p^\Phi(\omega)}$ is localized, namely its evaluation can be
restricted on a bounded real-space domain.
%The localization character of fluctuation states can be assessed by following the same argument discussed in section \ref{SEopenSystem} for the eigenstates of the SE. In this case 
%the source term depends also from the perturbation added to the system so this analysis applies only to the class 
%of perturbing field for which $\ket{\op\Phi\psi_p}$ is localized, when projected
%in the $\r$-representation. 
Even when such locality argument holds true, for $\omega > |\eps_p|$, 
the Helmholtz kernel behaves as a spherical wave and  gives raise to a $\ket{f^\Phi_p(\omega)}$ which is \emph{delocalized} in the entire real-space domain.
Such delocalization is an intrinsic consequence of the value of $\omega$, and does not depend on the particular choice of 
$\op \Phi$.
Only for $\omega < |\eps_p|$ the kernel contains an exponential damping factor and $\ket{f^\Phi_p(\omega)}$ exhibits a bound-state behavior in the long range, 
once again provided that the above locality arguments on $\ket{s_p^\Phi(\omega)}$ -- hence on $\op V'$ and $\Phi$ -- are valid.

Such locality property of the fluctuation state can be directly confirmed for static perturbations at $\omega=0$.
Let us consider a static perturbing operator $\op\Phi_s$ that can be restricted to a finite domain. 
Any local operator, like \emph{e.g.} a static electric field $\mathbf F$, with associated scalar potential $\Phi_s[\mathbf F](\r) = -\mathbf F \cdot \r$, would satisfy such property.
We consider the perturbed hamiltonian $\hnot[\dm_{\Phi_s}] + \op\Phi_s[\mathbf F]$, where $\dm_{\Phi_s}$ represents the GS density of the perturbed system, and assume that the 
perturbation strength ($|\mathbf F|$ in the above example) is sufficiently small to be consistent with a LR treatment of the problem. 
In this case, the occupied eigenfunction $\ket{\psi_p^{\Phi_s}}$ of the perturbed hamiltonian can be extracted with traditional GS techniques, 
as from the discussion around Eq.~\eqref{SEHelmholtz1}, they exhibit bound-state behavior.\footnote{The fulfilling of Eq.~\eqref{SEHelmholtz1} for energies below threshold
is not \textit{per se} a guarantee of a localized $\ket{\psi}$. With such equation we demonstrate that, \emph{in the case when} a localized
solution exists, this cannot be found for energies above threshold.}
\begin{figure}[t]
\includegraphics[scale=0.68]{CO_statPolvsBox.eps}
\caption{\label{co_alphaStatic}(Color online) Convergence of the statical polarizability (blue continuous line) and of the ground state permanent dipole (red dotted line) 
of the $CO$ molecule with respect to the size of the computational domain. The inset plot evidences the relative percentage convergence using the value associated to the largest 
box as reference.}
\end{figure}

From the linearized Schr\"odinger Equation of the perturbed problem, it is easy to see that the fluctuation state at $\omega=0$ can be expressed
in terms of the corresponding bound state of the perturbed Hamiltonian, i.e.
$\ket{f_p^{\Phi_s}(0)} \simeq \op Q_0 \ket{\psi_p^{\Phi_s}}$.
For any value of $p$, the locality of the $\ket{f_p^{\Phi_s}(0)}$ therefore directly stems from the
bound-state behavior of $\ket{\psi_p^{\Phi_s}}$. As $\omega=0$ is by hypothesis 
lower than any of the $|\eps_p|$, this is consistent with the above considerations.

Such observations enable us to express quantities like the static polarizability tensor $\alpha_{ij}$ from GS calculations of the perturbed Hamiltonian.
By expressing the induced dipole $\braket{\delta \op{\mathbf r}}_{\Phi_s}$ in the form Eq.~\eqref{LinearResponseFunctDef1}, and identifying the static response density as $\dm'_{\Phi_s}(0) = \dm_{\Phi_s} -\dmnot $, we have
\be \label{staticalpha}
\alpha_{ij} = 
-\frac{1}{F_j} \left(\trace{\op r_i \dm_{\Phi_s[F_j]}} - d^{\text{(GS)}}_i \right)\;,
\ee
% A direct confirmation of the localization character of fluctuation states at $\omega=0$ can be achieved through a ground-state calculation. Indeed, $\ket{f_p(\omega)}$ are nothing but the 
% perturbed KS eigenstates projected out from the domain of $\dmnot$, so for a constant perturbation $\op\Phi_s$ it turns out that $\ket{f_p(0)} = \op Q_0\ket{\psi_p^{\Phi_s}}$. The perturbed KS 
% states $\ket{\psi_p^{\Phi_s}}$ can be computed (beyond LR) by performing a DFT ground-state computation in which the constant perturbing field is added to the external potential and the 
% spatial extension of the fluctuation states can be directly measured.
where $\mathbf d^\text{(GS)}$ is the ground state dipole.
In Fig.~\eqref{co_alphaStatic} illustrate this procedure for a $CO$ molecule 
perturbed by a static electric field. The local character of the fluctuation states at $\omega=0$ is 
probed by verifying the convergence of the statical polarizability versus 
the size of the computational domain used to express the fluctuation states. The convergence rate is compared with the 
one of the static dipole of the molecule computed at zero external field. 
The plots evidence that the typical sizes of $\ket{f^\Phi_p(0)}$ are analogous to the ones of the unperturbed KS orbitals.

@@

The analysis described above has shown that the building blocks of the response density $\{\ket{f_p(\omega)}\}$ 
are susceptible to exhibit bound-state long-range behavior \emph{only} when $\omega < |\eps_p|$, 
while they behave as unbound states for values of $\omega$ above the threshold levels.

This fact has interesting implication in the evaluation of LR quantities. Indeed, writing the response 
density in terms of fluctuation states implies that 
\be\lb{LinearResponseFunctDef2}
\braket{\delta\op O}_\Phi(\omega) = 2\mathrm{Re} \left(\sum_p \bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}\right) \;.
\ee
%so the computation of the linear response functional passes through the evaluation of the scalar products indicated in \eqref{LinearResponseFunctDef2}.
Let us suppose that also the observable of interest $\op O$ exhibit the locality properties of the potentials $ \op V$, $\op \Phi$, $\op V'$.
Once again, it would be enough to consider a local operator for this condition to be met.
We observe that there exists two 
distinct regimes, on the basis of value of $\omega$. 
The first one, which we will refer to as the \emph{below threshold} regime, is realized when
$\omega$ is lower than the ionization potential $|\eps_h|$.
In this case all the fluctuation states are below their threshold level and equation \eqref{LinearResponseFunctDef2} 
is expressed in terms of genuinely localized 
quantities. The quantities $\braket{\psi_p \op O| f_p^\Phi(\omega)}$, 
being scalar products of bound-state-like objects, may thus be evaluated with computational setups that
are similar to the ones usually employed for GS calculations.
This consideration proves why, when a given computer code employs established GS numerical techniques, 
\emph{it is much easier to converge LR quantities for $\omega$ below IP}.
The convergence in this regime is ``strong'', in the sense that the basis functions employed enable to express 
with arbitrary precision \emph{both} the states $\ket{\op O \psi_p}$ and $\ket{f_p^\Phi(\omega)}$.
This is of course true even when the above states are only implicitly expressed by the employed LR treatment,
 as e.g. in the case of Krylov spaces generated from the states $\ket{\op\Phi(\omega) \psi_p}$ \cite{baroni2006,baroni2008,linlinKPM}.

@@

\subsection{Benzene}

The molecule is oriented in the $xy$ plain and the $x$-dimension of the box is equal to 15.0, 18.2 and 19.7 \AA.
220 states have been computed in each setup and the corresponding
energy reange for the unoccupied states is 16.9, 11.6 and 9.7 eV.

>>>>

%\title{On the analytic structure of the linear response susceptibility in open systems}
\title{Locality and Computational Reliability of Linear Response Calculations for Molecular Systems}% Localization properties, observable character and computational reliability of the linear response in open systems}
\author{Marco D'Alessandro}%
\affiliation{Istituto di Struttura della Materia-CNR (ISM-CNR), Via del Fosso del Cavaliere 100, 00133 Roma, Italia}
\author{Luigi Genovese}
\affiliation{Laboratoire de Simulation Atomistique (LSim), INAC, CEA-UGA, 17 Av. des Martyrs,
38054 Grenoble, France}
\date{\today}

@@

\section{Introduction}
Since the days of their foundations, disciplines like Computational Physics or Quantum Chemistry always had to deal with the problem of the 
computational reliability of the results. More specifically, a computational result may be considered reliable for various reasons.
It may be, for example, in agreement with some previous observation or measurement that the employed computational model is susceptible to reproduce.
The model itself may in this way be classified as ``accurate''; it can then be employed to extract, \textit{in silico},
other quantities that might be then used to \emph{predict} or \emph{interpret} experiments.
Another concept of reliability is the concept of  ``precision'', which is more focused to the actual results
that a given computational model provides for a specific set of quantities, regardless of their potential agreement with 
experimental data. The aim of precise computational approach is to reduce the \emph{computational} uncertainties
of quantities extracted out of a well-defined model, and provide reference results, to which other computer codes
employing the same model can be compared.

@@

The problem is even more stringent for calculations that refer to quantities \emph{beyond} ground state, like
for instance the time-dependent density functional theory (TDDFT) \cite{casida1995,runge1984,onida2002},
from which quantities like optical excitations or polarizability tensors
of a given system may be studied and put in relation with experimental data.
Here also, a plethora of computational approaches exist, that employ different numerical formalism and
basis sets, and it will be of great importance to identify quantities that can be ``measured'' in a given model, to assess
and quantify their computational uncertainty.

In this paper, we focus our attention to Linear-Response treatments of \emph{molecular} systems, 
in particular referring to LR-TDDFT calculation. We will investigate the impact that 
\emph{open} (i.e. Isolated) Boundary Conditions (BC) that have to be imposed to the system will have on the computational evaluation of LR quantities.
In order to do this, we perform analytic considerations on the equations beyond LR treatment.
Such considerations will enable us to distinguish quantities that \textit{a priori} can be compared among different treatments - and therefore tested for reproducibility - 
from others that would \emph{explicitly} depend of the employed computational treatment. 
In particular, we will discuss the interplay between reproducible LR quantities and the \emph{locality} of the related objects.
% In order to do this, we perform analytic considerations on the equations beyond LR treatment,
% which will help us identifying how to distinguish -- \textit{a priori} -- quantities that are susceptible to be 
% compared among different treatments, as they are \emph{intrinsically} associated to localized objects,
% from quantities that \emph{explicitly} depend of the the employed computational treatment.

The paper is organized as follows.
Initially we will inspect how the question of reproducibility may be expressed 
for ground state calculations of molecules. 
Following these guidelines, we will then move to LR equations,
by first considering the equations of motion of the response density operator, given
the application of a perturbing field.
Then we will consider the behaviour of the ``free oscillations'', i.e. the excited states
of the molecule.
Our considerations will also reveal useful for the understanding of the \emph{analytic structure} of LR quantities like
the linear susceptibility of the system (i.e. the reducible polarizability).
We support our considerations with numerical results on simple LR-TDDFT quantities.
For readability purposes, the details of each calculation are explained in the Appendix.

Although we refer, as anticipated, to TDDFT models, such considerations might also be transposed to Many-Body Perturbation Theory
approaches, in the context of Quasi-Particle equations.

@@

On the other hand, positive energy solutions behave differently. 
Even though the potential is localized, their positive energy are 
 spherical waves and the convolution with the source term provides a delocalized wavefunction $\psi(\r)$ in the whole space.
As they cannot be normalized they belong to the essential spectrum of the Hamiltonian operator and form a \emph{continuum} of states.
Outside of the support of the $\ket{\op V \psi}$ the solutions of Eq.~\eqref{SEHelmholtz1} can be put in bijection with the eigenstates of the 
free-particle Hamiltonian $\op H_s$. For positive energies, the quantum numbers of the eigenstates are determined by the Laplacian operator, 
regardless of the particular features of the potential.

It is easy to see from these concepts that a computational treatment that is tailored for the discretization of bound states
will not be as effective for continuum states, due to the different long-range asymptotic behavior - which in turn is a clear consequence of the 
BC of the problem. In a numerical treatment, where we discretize the problem with a \emph{finite}
number of degrees of freedom, we will only have access to a \emph{pseudo}-continuum of states,
whose energy values and density of states will depend on the representation of the kinetic operator
in the basis set. As a consequence, the set of eigenstates with $\epsilon > 0$ expressed in a given computational setup depends on the features of the employed basis set.

A direct numerical evidence of this problem has been discussed in \cite{boffi2016}, where the authors analyze the effects of a finite basis set on the 
occupied and virtual orbitals of a molecular system and establish a clear correspondence between the orbital localization character, 
and the independence of its energy with respect to the basis set. 
% It is shown that (localized) orbitals with a bound-state behavior in the long range constitute a genuine discrete set and a finite basis is 
% able to provide a reliable description of these states if the sampled domain contains their natural sizes. On the contrary, unbound (delocalized) orbitals form a continuum of states in the 
% true system with open boundaries. For these class of objects the presence of a finite computational domain has a deeper impact since introduces a fictitious discretization of the energy levels. 
Furthermore, it is clearly shown that the energies of the unbound orbitals are not stable with respect to parameters like the size of the computational domain.

\section{Fluctuation states. A localization argument for the response density}
It is evident that the arguments presented above have a limited interest for the study
of Ground State quantities of molecules. By definition, all these quantities can be expressed as a 
functional which \emph{entirely} depends on bound states. As an illustration it is sufficient to recall
the Hohenberg-Kohn theorem that states that the Ground-State (GS) energy is a functional of the GS
charge density $\dmnot$. For molecular systems that are electronically stable (i.e. no anionic or metastable configurations),
such quantity is localized, in the sense of the above section. Stated otherwise, 
\emph{there is no need to express efficiently any portion of the continuum spectrum} to have a reliable GS treatment of molecules.

@@

The response density satisfies an equation of motion written in the form of a quantum Liouville operator (for example see \cite{baroni2008})
\be\lb{LiouvillianRhopomegaDef1}
\left(\omega - \Liouv\right) \dm'_\Phi(\omega) =  \commutator{\op\Phi(\omega)}{\dmnot} \;.
\ee
The Liouvillian superoperator $\Liouv$ for self-consistent systems is expressed as the sum of the unperturbed part plus a coupling term. The action of these 
terms reads, respectively
\be\lb{LiouZeroDef1}
\Liouv \op O \equiv \left(\Liouvnot + \coupl \right) \op O = \commutator{\hnot}{\op O} +\commutator{\op V'[\op O]}{\dmnot}  \;,
\ee
where $\op V'[\op O] \equiv \int \dd \r \dd \r'\frac{\delta \op V[\dmnot]}{\delta \rho(\r,\r')} O(\r,\r')$ encodes the response of the $\dm$-dependent potential to a modification of the density operator.
The inspection of the linear order time evolution of the eigenstates of $\hnot$ evidence that the response density acts as a transition 
operator, linking the occupied and empty subspaces of $\hnot$.
An operator $\op O_\perp$ with this feature satisfy the \emph{transverse} condition  
\be\lb{RhopTransverseDef1}
\op O_\perp \equiv 
\dmnot\op O \op Q_0 + \op Q_0 \op O\dmnot \;,
\ee
where $\op Q_0=\identity-\dmnot$ is the projector in the subspace orthogonal to the occupied states of $\hnot$. 
Thus the condition $\dm'_\Phi = \dm_{\perp}$ holds.

We can therefore introduce an explicit representation of the 
response density parametrized as
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'_\Phi(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p^\Phi(-\omega)} + \ketbra{f_p^\Phi(\omega)}{\psi_p}\right) \;.
\ee
Here we have introduced the set of $\omega$-dependent \emph{fluctuation states} 
$\ket{f_p^\Phi(\omega)} = \dm'_\Phi(\omega) \ket{\psi_p}$, that have values in the unoccupied subspace of $\hnot$. 
An analysis of the Liouville equation \eqref{LiouvillianRhopomegaDef1} for the response density written in this fashion evidences that the equations of motion of fluctuation states (FS) are 
written as a modified Sternheimer equation~\cite{mahan1980,giustino2012,giustino2014}
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p^\Phi(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'[\dm'_\Phi](\omega))\ket{\psi_p} \;,
\ee
and we observe that the first-order Hamiltonian contains, 
apart from the perturbing field, a further term $\op V'[\dm'_\Phi]$ due to the density-dependence of $\hnot$.
% The interesting aspect of these equations in the perspective of the present analysis is that they allow us to present a localization argument for the fluctuation states using a procedure  
% analogous to the one depicted in section \ref{SEopenSystem}. So we split the unperturbed Hamiltonian as the sum of the kinetic $\op T$ and potential $\op V$ term and express the solution of 
% \eqref{fluctuationStateEqMotion1} in the form of a self-consistent inhomogeneous Helmholtz equation
It is easy to see that the solution of equation \eqref{fluctuationStateEqMotion1} can be casted in the form of
a set of scattering problems:
%a self-consistent inhomogeneous Helmholtz equation
\be\lb{rhoPrimeFluctuationStateDef2}
\ket{f_p^\Phi(\omega)} = \op G_H(\omega+\epsilon_p)\left(\op V\ket{f_p^\Phi(\omega)} + \ket{s_p^\Phi(\omega)} \right)\;,
\ee
with the state%:here in this case the source term contains further addend apart from the unperturbed potential and reads:
$
\ket{s_p^\Phi(\omega)} = \op Q_0(\op V'[\dm'_\Phi](\omega) + \op \Phi(\omega))\ket{\psi_p} 
$ as a source term.

It can be seen that for each choice of $p$, the value $\omega = |\eps_p|$ represents the threshold level 
that governs the localization properties of the FS.
As a \emph{stringent} hypothesis, let us now assume that the source term $\ket{s_p^\Phi(\omega)}$ is localized, \textit{i.e.} its evaluation can be
restricted on a bounded real-space domain.
%The localization character of fluctuation states can be assessed by following the same argument discussed in section \ref{SEopenSystem} for the eigenstates of the SE. In this case 
%the source term depends also from the perturbation added to the system so this analysis applies only to the class 
%of perturbing field for which $\ket{\op\Phi\psi_p}$ is localized, when projected
%in the $\r$-representation. 
Even when such locality argument holds true, for $\omega > |\eps_p|$, 
the Helmholtz kernel behaves as a spherical wave and  gives raise to a $\ket{f^\Phi_p(\omega)}$ which is \emph{delocalized} in the entire real-space domain.
Such delocalization is an intrinsic consequence of the value of $\omega$, and does not depend on the particular choice of 
$\op \Phi$.
Only for $\omega < |\eps_p|$ the kernel contains an exponential damping factor and $\ket{f^\Phi_p(\omega)}$ exhibits a bound-state behavior in the long range, 
once again provided that the above locality arguments on $\ket{s_p^\Phi(\omega)}$ -- hence on $\op V'$ and $\Phi$ -- are valid.

Such locality property of the fluctuation state can be directly confirmed for static perturbations at $\omega=0$.
Let us consider a static perturbing operator $\op\Phi_s$ that can be restricted to a finite domain. 
Any local operator, like \emph{e.g.} a static electric field $\mathbf F$, with associated scalar potential $\Phi_s[\mathbf F](\r) = -\mathbf F \cdot \r$, would satisfy such property.
We consider the perturbed hamiltonian $\hnot[\dm_{\Phi_s}] + \op\Phi_s[\mathbf F]$, where $\dm_{\Phi_s}$ represents the GS density of the perturbed system, and assume that the 
perturbation strength ($|\mathbf F|$ in the above example) is sufficiently small to be consistent with a LR treatment of the problem. 
In this case, the occupied eigenfunction $\ket{\psi_p^{\Phi_s}}$ of the perturbed hamiltonian can be extracted with traditional GS techniques, 
as from the discussion around Eq.~\eqref{SEHelmholtz1}, they exhibit bound-state behavior.
%\footnote{The fulfilling of Eq.~\eqref{SEHelmholtz1} for energies below threshold
% is not \textit{per se} a guarantee of a localized $\ket{\psi}$. With such equation we demonstrate that, \emph{in the case when} a localized
% solution exists, this cannot be found for energies above threshold.}
\begin{figure}[t]
\includegraphics[scale=0.68]{CO_statPolvsBox.eps}
\caption{\label{co_alphaStatic}(Color online) Convergence of the statical polarizability $\alpha_{zz}$ (blue continuous line) and of the ground state dipole $d_z$ (red dotted line) 
of $CO$ molecule as a function of the size of the computational domain. The inset evidences the relative convergence (in percent) of the two quantities w.r.t. the corresponding value associated to the largest 
box.}
\end{figure}

From the linearized Schr\"odinger Equation of the perturbed problem, it is easy to see that the FS at $\omega=0$ can be expressed
in terms of the corresponding bound state of the perturbed Hamiltonian, i.e.
$\ket{f_p^{\Phi_s}(0)} \simeq \op Q_0 \ket{\psi_p^{\Phi_s}}$.
For any value of $p$, the locality of the $\ket{f_p^{\Phi_s}(0)}$ therefore directly stems from the
bound-state behavior of $\ket{\psi_p^{\Phi_s}}$. As $\omega=0$ is by hypothesis 
lower than any of the $|\eps_p|$, this is consistent with the above considerations.

Thanks to such observations one can express quantities like the static polarizability tensor $\alpha_{ij}$ (see e.g.\cite{DebElecField}) from GS calculations of the perturbed Hamiltonian.
By expressing the induced dipole $\braket{\delta \op{\mathbf r}}_{\Phi_s}$ in the form Eq.~\eqref{LinearResponseFunctDef1}, and identifying the static response density as $\dm'_{\Phi_s}(0) = \dm_{\Phi_s} -\dmnot $, we have
\be \label{staticalpha}
\alpha_{ij} = 
-\frac{1}{F_j} \left(\trace{\op r_i \dm_{\Phi_s[F_j]}} - d^{\text{(GS)}}_i \right)\;,
\ee
% A direct confirmation of the localization character of fluctuation states at $\omega=0$ can be achieved through a ground-state calculation. Indeed, $\ket{f_p(\omega)}$ are nothing but the 
% perturbed KS eigenstates projected out from the domain of $\dmnot$, so for a constant perturbation $\op\Phi_s$ it turns out that $\ket{f_p(0)} = \op Q_0\ket{\psi_p^{\Phi_s}}$. The perturbed KS 
% states $\ket{\psi_p^{\Phi_s}}$ can be computed (beyond LR) by performing a DFT ground-state computation in which the constant perturbing field is added to the external potential and the 
% spatial extension of the fluctuation states can be directly measured.
where $\mathbf d^\text{(GS)}$ is the ground state dipole.
Fig.~\eqref{co_alphaStatic} illustrates this procedure for a $CO$ molecule 
perturbed by a static electric field. The local character of the fluctuation states at $\omega=0$ is 
probed by verifying the convergence of the statical polarizability versus 
the size of the computational domain used to express the fluctuation states. The convergence rate is compared with the 
one of the static dipole of the molecule computed at zero external field. 
The results evidence that the typical sizes of $\ket{f^\Phi_p(0)}$ are analogous to the ones of the unperturbed KS orbitals, which is a direct proof of the bound-state behaviour of the FS.

@@

The analysis described above has shown that the building blocks of the response density $\{\ket{f_p(\omega)}\}$ 
are susceptible to exhibit bound-state long-range behavior \emph{only} when $\omega < |\eps_p|$, 
while they behave as unbound states for values of $\omega$ above the threshold levels.
This fact has interesting implications in the evaluation of LR quantities. Indeed, writing the response 
density in terms of FS implies that 
\be\lb{LinearResponseFunctDef2}
\braket{\delta\op O}_\Phi(\omega) = 2\mathrm{Re} \left(\sum_p \bra{\psi_p}\op O\ket{f_p^\Phi(\omega)}\right) \;.
\ee
%so the computation of the linear response functional passes through the evaluation of the scalar products indicated in \eqref{LinearResponseFunctDef2}.
Let us suppose that also the observable of interest $\op O$ exhibits the locality properties introduced for the potentials $ \op V$, $\op \Phi$ and $\op V'$.
Once again, it would be enough to consider a local operator for this condition to be met.

We observe that there exists two 
distinct regimes, on the basis of value of $\omega$. 
The first one, which we will refer to as the \emph{below threshold} regime, is realized when
$\omega$ is lower than the ionization potential $|\eps_h|$.
In this case all the FS are below their threshold level and equation \eqref{LinearResponseFunctDef2} 
is expressed in terms of genuinely localized 
quantities. At these frequencies, the linear response functional $\braket{\delta\op O}_\Phi(\omega)$
%The quantities $\braket{\psi_p \op O| f_p^\Phi(\omega)}$, 
%being scalar products of bound-state-like objects, 
may thus be evaluated with computational setups that
are similar to the ones usually employed for GS calculations.
The convergence in this regime is ``strong'', in the sense that the basis functions employed enables us to express 
with arbitrary precision \emph{both} the states $\ket{\op O \psi_p}$ and $\ket{f_p^\Phi(\omega)}$.
This is of course true even when the above states are only implicitly expressed by the employed LR treatment,
 as e.g. in the case of Krylov spaces generated from the states $\ket{\op\Phi(\omega) \psi_p}$ \cite{baroni2006,baroni2008,linlinKPM}.

@@

\subsection{Benzene}
The molecule is oriented in the $xy$ plain and the $x$-dimension of the box is equal to 15.0, 18.2 and 19.7 \AA.
220 states have been computed in each setup and the corresponding
energy reange for the unoccupied states is 16.9, 11.6 and 9.7 eV.
